---
title: "IntroDay2"
author: Lucy Liu
output:
  pdf_document: default
  html_document: 
    keep_md: true
---

#Summary Day 1

First for good practice lets clean our environment - we have made lots of variables with lots of names. This can get messy so for a fresh start lets delete them all.

##Variables
We learnt how a variable is just a thing that we have given a name to. They appear here in our environment.
```{r}
read.csv("cat.csv")
#this just outputs here in the console but it's not saved. 
#I can't use it again because I haven't named it.
```
```{r}
#But if I create a variable by giving this a name - 
cats <- read.csv("cat.csv") 
#it pops up here and I can use it again by referring to it by name.
```

##Functions
We also learnt about functions. R has lots of functions built in already. To use them you write the name of the function - and then ALWAYS brackets. Inside the brackets you usually have to give it some input. These inputs are called arguments.   
**%Whiteboard%**  
We can give the arguments in 2 ways. First we can specify what the argument is.
```{r}
rnorm(n=4, mean = 5, sd = 1)
```
If we do it like this, it doesn't matter which order we do this.
```{r}
rnorm(sd = 1, n=2, mean = 4)
```
It gives us what we want.  
We could also not give these specifications. What will R give us?
```{r}
rnorm(2,3,4)
```
If you do it this way - order matters! R expects the first number to be how many random numbers, the second to be the mean and the third to be the sd.  
This is the same for all the functions you use in R.  

##Making functions
We also learnt how to make our own functions. This helps us because if you wanted to do 10 things - instead of typing out all your 10 steps - you write a function that will do all your 10 steps. We made a function to turn fahrenheit to kelvin.

```{r, eval=FALSE, echo=TRUE}
#Kelvin = (Fahrenheit – 32) * 5/9 + 273.15
#instead of typing this out every time we want to do a conversion - 
#we made a function so we just have to write the function name and give it a temp.

F_to_K <- function(temp){
  
  kelvin <- ((temp - 32) * (5 / 9)) + 273.15
  return(kelvin)
}
#temp could be anything - we could change it to bob - 
#but we just need to change it down here as well.

F_to_K()
#We give it a number. When we look back at where we created the function - 
#we can see that this is what the function is doing to the number. 
```
**%Whiteboard%**  
In the round brackets you say what you want people to input when they use the function. So we want people to give you 1 thing so here we put 1 name - this could be any name - we are just naming it so we can refer to it down here. Here we say what we want to do with the input. When we use it - write function name and brackets.

Any questions? 

#Data structures & types
We also learnt about data structures - this is a format for storing your data - think of it as a container for your data but the container has specifications for how data is ordered and what you should put where.

**%Whiteboard%** 
Vectors - 1D and all are labelled as the same data type.
Dataframe - 2D, like your excel spreadsheet - all columns are the same data type.

We learnt about 2 others but you won't use them very often.

##Data types
Every bit of data in R is labelled as 1 data type. This doesn't change your data in any way. E.g.
```{r}
cats$weight
typeof(cats$weight)
as.character(cats$weight)
```
What do you think will happen here?
```{r}
cats$likes_string
as.integer(cats$likes_string)
```
The label is just for R to know what it can and can't do with that bit of data - you can log a number but not a character.  
**%Whiteboard%** These are the basic data types.

###Coercion
If we have different data types in a vector or column - e.g. characters and numbers - R will label the whole vector as one data type - but it will do it in a way so that no data is lost. What will happen here? What will our vector look like?
```{r}
vector1 <- c(2,4,"dan")
#remember dan can't be a number
typeof(vector1)
```
```{r}
vector2 <- c(TRUE, FALSE, 4)
vector3 <- c(TRUE, FALSE, "i like cats")
```
Data exactly the same.

###NA and factors
A useful thing to know is what to do when your data has a whole column of numbers and someone has put xxx or missing when that bit of data is missing. %%Use new csv%%  
What do you think this data will look like?
```{r}
catsDay2 <- read.csv("catDay2.csv")
catsDay2
```
Data is not changed but labelling has.

###Factors
What datatype is this?
```{r, error=TRUE}
str(catsDay2$weight)
```
Remember the factors? It is another data type and it is used for labelling for categorical data. A funny quirk of R is that any column that has words/characters in it - R will read it in as a factor.

We can see here that it says there are 4 levels - 4 possible categories. And R thinks of each bit of data in this column as being one of these categories.

What we want is for R to know that this column should be numbers and when it sees the word missing - it is actually a NA value. We can do this easily with an option in the function.
```{r}
catsDay2 <- read.csv("catDay2.csv", na.strings = "missing")
```
We can see now that the missing 

***
CHALLENGE: Change the function below such that the word missing AND Missing are both recorded as NA values in R. 

catsDay2 <- read.csv("cat.csv", na.strings = "missing")  

***

```{r}
catsDay2 <- read.csv("catDay2.csv", na.strings = c("missing","Missing"))
```

Note that commas are special in R. They are used to separate arguments/inputs. 

###NA
The good think about R knowing that these values as NAs is you can do things like this. Say you want to exclude all rows that have an NA somewhere in it.
```{r}
na.omit(catsDay2) 
```
The other thing to note about NA is that they are contagious. **%Whiteboard%**  
```{r}
log(NA)
```
This makes sense. You can't take the log of a value you don't know the value of.  
What do you think will happen here?
```{r}
vector1 <- c(1,2,3,NA)
mean(vector1)
```
This gives NA because R says I don't know what the mean is because I don't know what this value is. You can tell it to ignore the NA value though.
```{r}
mean(vector1, na.rm = TRUE)
```
A lot of functions have this option to ignore NAs. log, sum etc

##Subsetting
To get a portion of your vector or data frame.  
**%Whiteboard%** You use square brackets - this is important.  
For vectors you put 1 thing - which element/s you want.  
For dfs you put 2 things - which row/s you want and which column/s you want. Commas are special!!!

This is my vector.
```{r}
vector1 <- 11:20
vector1
```
How do I get the 6th element?
```{r}
vector1[6]
```
How do I get the 1st, 5th and 9th elements?
```{r}
vector1[c(1,5,9)]
```
Let's look at a dataframe.
```{r}
titanic <- read.csv("https://goo.gl/4Gqsnz")
```

How do I get the first 10 rows and the 1st and 2nd columns?
```{r}
titanic[1:10, c(1,2)]
```
What if I don't know the number of the element I want - I just know I want the first 10 rows and the name column?  
This would be ideal.
```{r, error=TRUE}
titanic[1:10,Name]
```
The square brackets don't work this way though. It can't search. I can give it either numbers or I can use another function to search - this function will give TRUEs and FALSEs and then I can give this to the square bracket.  
I can get the column names like this.
```{r}
colnames(titanic)
```
See how this is in the same order as the dataframe?  
Now I need to search this for the Name column. I can do this with the %in% function we used yesterday. The notation we used was this:
```{r}
colnames(titanic)%in%c("Name")
```
This function searches colnames and finds which one is equal to Names. It returns this T and F. Remember I can only put numbers or T and F into the square brackets. **%Whiteboard%** It puts this into the second part - to know which columns we want.  
What will this give us?
```{r}
titanic[1:10,colnames(titanic)%in%c("Name")]
```

***
CHALLENGE: Change the code below so that we get the names NOT of the first 10 rows but of the rows where the cabin == C85. (Remember we need to use == when comparing things)

Hint use titanic$Cabin  

***

```{r}
titanic[titanic$Cabin==c("C85"),colnames(titanic)%in%c("Name")]
```

Any questions?

**Take 10 min break**

#Control flow 
Introduce Miriam.

**Break or Lunch** 

#dplyr
We've learnt how to manipulate data by subsetting it to get groups that we want yesterday. Now we are going to learn about a package called dplyr which was designed for data manipulation and is more intuitive to use than the standard R functions. Lets download and load the package.  

```{r}
#install.packages('dplyr') #downloads and saves to your computer. Only have to do this once.
library(dplyr) #do this for each new session of R.
```
We will use the titanic data set again. I'll load it again in case anyone has deleted it or weren't here on day 1.
```{r}
titanic <- read.csv("https://goo.gl/4Gqsnz")
titanic <- na.omit(titanic)
```
##Select
Select function lets you select certain columns. You use it like this: 
```{r, eval=FALSE, echo=TRUE}
titanic %>%
  select(Name, Sex, Survived)
```
%>% is called the pipe symbol and you can think of it like a pipe sending your data to the next function. So the titanic data is going to the fuction selection, which selects the columns Name, Sex and Survived.  
If you wanted to save this new dataframe you can do this:
```{r}
new_df <- titanic %>%
  select(Name, Sex, Survived)
```

##Filter
This function filters the data. What will this do?
```{r, eval=FALSE, echo=TRUE}
titanic %>%
    filter(Sex=="female") %>%
    select(Name,Sex,Survived)
```
Here we have filtered the data to include only female passengers - think of this as removing the rows of men, then we send the data to the select function to select only the columns Name, Sex and Survived.

***
CHALLENGE: Write a single command (which can span multiple lines and includes pipes) that will produce a dataframe that has the values for Age, SibSp and Fare for males only. How many rows does your dataframe have and why?  

***

#group_by and summarise
The next functions we'll learn about are group_by and summarise (write on whiteboard). These are great when you use them together. What will this do?
```{r}
survived <- titanic %>%
              group_by(Sex) %>%
              summarise(mean_Survived=mean(Survived))
```

```{r}
str(survived)
```
Group_by gives you the same dataframe you had before - the rows are just grouped by sex. When you add summarise - it adds a new column that gives the mean of the survived column for each group.  

***
CHALLENGE: Calculate the average Survived value per Pclass and Sex. Which combination of Pclass and Sex had the highest Survived value and which combination had the lowest?  

***

```{r}
titanic %>%
  group_by(Pclass, Sex) %>%
  summarise(mean_Survived = mean(Survived))
```


##mutate
mutate creates a new column.
```{r}
Survived_by_Sex_Status <- titanic %>%
    mutate(Status=Age*Pclass) %>%
    group_by(Sex) %>%
    summarize(mean_Status=mean(Status),
              sd_Status=sd(Status))
Survived_by_Sex_Status
```
  
This next challenge is really tricky so you can go to lunch after doing it.  

***
CHALLENGE: Calculate the average Survived value for female passengers from each Pclass group. Then arrange the classes in descending order. 

Hint: Use the dplyr function arrange() and desc(). Look at the help!  

***

```{r}
titanic %>%
  filter(Sex == "female") %>%
  group_by(Pclass) %>%
  summarise(mean_survived = mean(Survived)) %>%
  arrange(desc(Pclass))

```


**Break**

#dplyr and ggplot
You can use dplyr and ggplot together!  
We are going to use data about Melbourne house prices! I definitely can't afford anything here.
```{r}
houses <- read.csv("https://docs.google.com/uc?id=1wdSyZZ3U4tqMfgCKin7ogpTKaQ9gAc4t&export=download")
houses <- na.omit(houses)
```

```{r}
#install.packages("ggplot2")
library(ggplot2)
```

##Scatter plot
Lets just start plotting some things. Does not look good. I want to look at this bit of data here and I want to exclude this. You can do this with coord_cartesian().
```{r}
ggplot(houses) +
  geom_point(aes(x = Landsize, y = Price)) + 
  coord_cartesian(xlim = c(0,2000))
```
  
***
CHALLENGE: Make a scatter plot of Landsize vs YearBuilt. Change the coordinates of the y and x axis until you can visualise the data better.

***

```{r}
ggplot(houses) +
  geom_point(aes(y = Landsize, x = YearBuilt)) +
  coord_cartesian(xlim = c(1700,2150), ylim = c(0,10000))
```

##Boxplot
```{r}
ggplot(houses) + 
  geom_boxplot(aes(y = Price, x = Type))
```

***
CHALLENGE: Make a boxplot of Price for each CouncilArea

***

We can't really read the labels. To fix this we can switch the axis.
```{r}
ggplot(houses) + 
  geom_boxplot(aes(y = Price, x = CouncilArea)) +
  coord_flip()
```
So this is getting a bit messy. We are only interested in some areas anyway. We can combine dplyr (the select and filter we were using before) with ggplot.  
We start filtering using
```{r, eval=FALSE, echo=TRUE}
houses %>%
  filter(CouncilArea %in% c("Maribyrnong", "Melbourne"))
```
Should I use == or %in% ?????

If you remember from subsetting. What will this give me?
```{r}
vector1 <- 1:10
vector1 %in% 0:5
```
What will this give me?
```{r}
vector1
0:5
vector1 == 0:5
```
Back to our plot. Note how I am sending the data - think of data flowing through these pipes - to ggplot. I dont have to specify what data I am using because it is being sent in from here.
```{r}
houses %>%
  filter(CouncilArea %in% c("Maribyrnong", "Melbourne")) %>%
  ggplot() +
  geom_boxplot(aes(y = Price, x = CouncilArea))
```

We can also do a side by side boxplot. Say for every CouncilArea, we want to know the price for each type of house we just have to add fill = Type to aesthetics.
```{r}
houses %>%
  filter(CouncilArea %in% c("Maribyrnong", "Melbourne")) %>%
  ggplot() +
  geom_boxplot(aes(y = Price, x = CouncilArea, fill = Type))
```
  
```{r}
houses %>%
  filter(CouncilArea %in% c("Maribyrnong", "Melbourne")) %>%
  ggplot() +
  geom_boxplot(aes(y = Price, x = CouncilArea, fill = Type)) +
  scale_fill_manual(values = c("pink", "green", "red"), name = "Type of house", labels = c("house","townhouse","unit"))
```

***
CHALLENGE: Pick 3 CouncilAreas and make your own side by side boxplot. Use scale_fill_manual() to change the colours of the boxes and the labels.

***

##Facet grid

```{r}
houses %>%
  filter(CouncilArea %in% c("Maribyrnong", "Melbourne")) %>%
  ggplot() +
  geom_boxplot(aes(y = Price, x = Type)) +
  facet_grid(.~CouncilArea) 
```

Lets make this look better. First we can add titles.
```{r}
houses %>%
  filter(CouncilArea %in% c("Maribyrnong", "Melbourne")) %>%
  ggplot() +
  geom_boxplot(aes(y = Price, x = Type)) +
  scale_x_discrete(labels = c("House", "Townhouse", "Unit")) +
  facet_grid(.~CouncilArea) +
  labs(title = "title", y = "So expensve", x = "types")
```

***
Challenge: Change the code so that we have price in million on the y axis. Then change x axis label so that it tells us the price is in millions.

Hint: Make a new column with mutate that is the price in million.

***

#Saving plots
So I like this plot and I want to save it. Show how to save plots in plots using Export.  
Another way to save is to use pdf(). This is good if you are making 10 plots and you don't want to click through for each one. 
```{r, eval=FALSE, echo=TRUE}
pdf("myPlot.pdf", width = 10, height = 8)

houses %>%
  ggplot() +
  geom_boxplot(aes(y = Price, x = Type)) 

dev.off
```

##Bar plots
Let's make bar plots now! 
**%Whiteboard%** I want to plot mean price for each number of rooms. First I need to work out the mean price for each number of rooms. Which one of these would I use? 
```{r}
houses %>%
  group_by(Bedroom2) %>%
  summarise(meanPrice = mean(Price))
```
Lets plot this. Note that I send the data through to ggplot. 
```{r, eval=FALSE, echo=TRUE}
houses %>%
  group_by(Bedroom2) %>%
  summarise(meanPrice = mean(Price, na.rm = TRUE)) %>%
  ggplot() + 
    geom_bar(aes(y = meanPrice, x = Bedroom2))  
```
Why have we gotten this error??  
Let's look at a bar plot that doesn't give an error.
```{r}
ggplot(houses) +
  geom_bar(aes(x = Bedroom2))
```
What is this plot showing? y = number of houses that had 1 bedroom, 2bds....  
What this function has done, is automatically counted how many houses/rows have 1 bd, 2 bd...  
Notice how we only gave it 1 variable - because it's just counting up how many times there is a 1 bd house and putting the count herre - in the y axis.  
Lets look at our previous data again. We are not plotting how many - we want it to plot actually this number here. We can tell it to do so with stat = "identity"
```{r}
houses %>%
  group_by(Bedroom2) %>%
  summarise(meanPrice = mean(Price, na.rm = TRUE)) %>%
  ggplot() + 
    geom_bar(aes(y = meanPrice, x = Bedroom2), stat = "identity")  
```

So I only want to look at houses that have 5 bedrooms or less - I'm a poor PhD student. Remember should I use == or %in%  ??

```{r}
houses %>%
  filter(Bedroom2 %in% 0:5) %>%
  group_by(Bedroom2) %>%
  summarise(meanPrice = mean(Price, na.rm = TRUE)) %>%
  ggplot() + 
    geom_bar(aes(y = meanPrice, x = Bedroom2), stat = "identity") 

```
  
Okay back to our plot. I want the ticks here for each bedroom.
```{r}
houses %>%
  filter(Bedroom2 %in% 0:5) %>%
  group_by(Bedroom2) %>%
  summarise(meanPrice = mean(Price, na.rm = TRUE)) %>%
  ggplot() + 
    geom_bar(aes(y = meanPrice, x = Bedroom2), stat = "identity") +
    scale_x_continuous(breaks = 0:5, labels = 0:5)
#breaks mean the points at which the grid lines appear
```

***
CHALLENGE: Make a side by side bar plot so for every bedroom number you have average price for each house Type.

Hint: You will need to make a dataframe that shows mean price for each house type AND bedroom number then give this to ggplot.  
Use position = "dodge" in geom_bar() to get the bars to be side by side
Extra: Change the colours of the bars with scale_fill_manual()

***

```{r}
houses %>%
  filter(Bedroom2 %in% 0:5) %>%
  group_by(Bedroom2, Type) %>%
  summarise(meanPrice = mean(Price, na.rm = TRUE)) %>%
  ggplot() + 
    geom_bar(aes(y = meanPrice, x = Bedroom2, fill = Type), stat = "identity", position = "dodge") +
    scale_x_continuous(breaks = 0:5, labels = 0:5)
```

This is not very good looking but you can change the labels and the colours if you want.  
Now I want one of these plots for Brunswick. And I only want 1-3 bedrooms.
```{r}
houses %>%
  filter(Bedroom2 %in% 1:3 & Suburb %in% c("Brunswick")) %>%
  group_by(Bedroom2, Type) %>%
  summarise(meanPrice = mean(Price, na.rm = TRUE) / 1000000) %>%
  ggplot() + 
    geom_bar(aes(y = meanPrice, x = Bedroom2, fill = Type), stat = "identity", position = "dodge") + 
    labs(y = "price in millions") 
```
So that is 1 Suburb. What if I wanted to do this for a vector of Suburbs? I can write a for loop.

***
CHALLENGE: Write a for loop that will give you Price vs bedrooms (separated by house type) for 3 Suburbs. Add a title that states which Suburb the plot is for.
Note when using ggplot in a for loop you have to specify that you want to print the plot with print().  
i.e. print(
``to make plot``
)

***

```{r, eval=FALSE, echo=TRUE}
for (i in coolSuburbs){
print(
  houses %>%
  filter(Bedroom2 %in% 1:3 & Suburb %in% i) %>%
  group_by(Bedroom2, Type) %>%
  summarise(meanPrice = mean(Price, na.rm = TRUE) / 1000000) %>%
  ggplot() + 
    geom_bar(aes(y = meanPrice, x = Bedroom2, fill = Type), stat = "identity", position = "dodge") + 
    labs(y = "price in millions", title = i) 
  
)
}

```

You can save mulitple plots to one pdf file using pdf(). I want this plot for the suburbs I'm interested in. 














