---
title: "Lucy's ResPitch"
output:
  pdf_document: default
  html_document: default
---

We are going to be using ggplot. First load it so we can use it.
```{r}
library(ggplot2)
```

This line imports my data into R. Show R object.   
We are going to turn this into a great graph because it's difficult to take away any patterns by just looking at these numbers. 
```{r}
hp_data <- 
  read.csv("https://docs.google.com/uc?id=0B2jjnIpe-k4MMmgxRVAtamZpVkE&export=download")
```

We start by telling R we want to use ggplot. We do this by typing in the ggplot with (). 

Inside the round brackets, we tell ggplot what data we want to plot. We then run the code. We get a grey background. This is the base of our plot.
```{r}
ggplot(data = hp_data)
```
  
1. Next we want to add a plot. We will make a bar graph by 'adding' another layer. We do this by putting a plus sign at the end, this tells R that we want to add another layer and there will be more instructions on the next line.
2. The next 'layer' we will add is the graph. Different graphs are called "geoms" in ggplot. This is short for geometric object because a graph uses geometric objects to represent data. We want to add a bar plot so we will use "geom_col" - columns
3. Now we need to tell R more details on how to make our bar plot. We need to tell it what goes on the x and y axes. To do this, first we add aes()  - "aesthetics" - it is in here that we tell R how to make our plot. Within these round brackets we will tell R that on the X axis we want 'House' and on the y-axis we want "Percent", the percent of people that were sorted into that house

```{r}
ggplot(data = hp_data) +
  geom_col(aes(x = House, y = Percent))
```
  
With only 2 lines of code - "instructions to the computer" we have a decent plot. The great thing about R is that we can add features and customise this as much as we want!  
Lets make this fancier by adding a third variable! You will notice in the table that we also have the sex of people sorted into each house. Can we make a bar plot where we can show the sex of people sorted into each house? Yes! Very easily by adding just 2 words! So within aes() - where we tell R about the details of the plot we will add fill = sex, we want the fill - the colour inside the bars - to be colours according to sex.

```{r}
ggplot(data = hp_data) +
  geom_col(aes(x = House, y = Percent, fill = Sex))
```
  
This is good but I don't like the way that bars are stack on top of each other. It's ugly and it's hard to compare.  
We can change this easily by adding another 2 words. We will add them at the end, position = dodge This tells R that we want the bars to be next to each other.
```{r}
ggplot(data = hp_data) +
  geom_col(aes(x = House, y = Percent, fill = Sex), position = "dodge")
```
Now we can see something interesting. There are more males in Gryffindor and Slytherin and more females in Hufflepuff and Ravenclaw - read into that what you will....  
Now this is not publication ready. We want to add a title and maybe change the colours. This can be done really easily by "adding" layers! Lets add a title first.
1. First we add a plus - this tells R that we want to add another layer and to expect more code/instructions
2. Then we want to add labs() - this lets you add labels
3. Inside the brackets we can tell R what labels we want to add. title tells R we want to change the title, then we tell it what we want the title to be using = "" - adding our title in quotes here.

```{r}
ggplot(data = hp_data) +
  geom_col(aes(x = House, y = Percent, fill = Sex), position = "dodge") +
  labs(title = "Hufflepuff 4 life!") 
```

